Evaluation of Design Trade-offs for Adders\\ in Approximate Datapath

### 2015-01-07 WAP2015/wap2015.tex
WAPCO15 paper final version

\usepackage{algorithm}
\usepackage{array}
\usepackage{algpseudocode}
\usepackage{indentfirst}
\usepackage[pdftex]{graphicx}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{multirow}

@@

In this paper, we provide detailed evaluations of design trade-offs for adders, which are key arithmetic primitives for a broad range of applications. In order to provide the optimum design choice for adders, we examine several design constrains and options, such as accuracy, operating frequency, silicon area, different types of adder structure, as well as different forms of computer arithmetic for adder implementation. In comparison to the conventional design scenario that timing closure is achieved by truncating the operand word-length, we take a radical step forward by adopting an alternative design scenario, which is named ``overclockign scenario''. In this case, circuits are operated far beyond the conservation region with full precision, while timing violation can be tolerated. Previous work has shown that for many applications, it is preferable to follow the overclocking scenario because the worst case is triggered only by certain input patterns and hence it happens rarely~\cite{SKfccm13}. In addition to the adder structures that are created based on standard binary arithmetic, such as ripple carry adder and carry select adder, we also evaluate adder with ``online arithmetic''~\cite{Ercegovac_OnlineOverview}, which employs different style of data representation to eliminate carry propagation. We demonstrate experimentally on FPGAs that for limited area budgets, applying the overclocking approach to ripple carry adders can be more beneficial than using fast adders to achieve similar latency. With more relaxed area constraints, the online adder outperforms in terms of accuracy or performance. Potentially our study will provide circuit designers with guidance and extra options when balancing between various design trade-offs.

@@

For an $n$-bit RCA, it is composed of $n$ serial-connected full adders (FA) as shown in Fig.~\ref{Fig:rca}. Typically the maximum frequency of RCA is determined by the longest carry propagation. Under the assumption that the carry propagation delay of each FA is a constant value $\mu_i=\mu$, the critical path of the RCA is: $\mu_{RCA}=n\mu$. For the sampling period $T_S$, it follows that if $T_S\geqslant\mu_{RCA}$, correct result will always be sampled. Otherwise, intermediate result will be sampled and potentially generating overclocking errors. The mean value of overclocking error is modeled~\cite{SKfccm13} as given by (\ref{Eq:MeanError_RCA}), where coefficient $b$ is given in (\ref{Eq:b}) and it determines the maximum length of error-free carry propagation.
%
\begin{eqnarray}\label{Eq:MeanError_RCA}
    E_O=\left\{
        \begin{matrix}
            2^{-b}-2^{-n-1}, & \textrm{if $b\leq n$}\\
            0, & \textrm{otherwise}
        \end{matrix}
        \right.
\end{eqnarray}
\vspace{-1.5ex}
%
\begin{eqnarray}\label{Eq:b}
    b:=\left\lceil \frac{T_S}{\mu} \right\rceil=\left\lceil \frac{1}{\mu\cdot f_S}\right\rceil
\end{eqnarray}
\vspace{-1ex}
\normalsize

In the conventional scenario, a specific timing target can be met by truncating the word-length of RCA while timing violation is not permitted. This will result in truncation error for most data.  The mean value of truncation error is also modeled~\cite{SKfccm13} as given by~(\ref{Eq:MeanTruncError_RCA}), where parameters $k$ and $n$ denote the word-length of RCA before and after truncation, respectively. 
%
\begin{eqnarray}\label{Eq:MeanTruncError_RCA}
    E_O=\left\{
        \begin{matrix}
            2^{-n}-2^{-k}, & \textrm{if $n<k$}\\
            0, & \textrm{otherwise}
        \end{matrix}
        \right.
\end{eqnarray}
\vspace{-2ex}

@@

\subsection{Online Adder}\label{subsec:online_adder}
In addition to standard binary arithmetic, alternative forms of computer arithmetic was studied to boost performance. For example, adder with online arithmetic is also designed for low latency at the cost of extra silicon area. The structure of a digit-parallel online adder (OA) where all signals represented with SD numbers of digit set $\{-1,0,1\}$ is shown in Fig.~\ref{Fig:Radix2SD_adder}. The module ``3:2'' denotes a 3:2 compressor, which takes three inputs and generates two outputs, and is logically equivalent to a full adder (FA). A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As labeled in Fig.~\ref{Fig:Radix2SD_adder}, ideally the computation delay of this adder is only two FA delays for any operand word-length, at the expenses of one extra FA for each digit of operands. 
%
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.45\textwidth]{./figures/SDAdder.eps}
  \vspace{-1ex}
  \caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
  \label{Fig:Radix2SD_adder}
  \vspace{-2ex}
\end{figure}

@@

As can be seen in Fig.~\ref{Fig:max_wl_adder}, for a relatively relaxed frequency requirement (e.g. 470 MHz), both CSA and OA can be implemented with greater word-lengths than RCA. For CSA, this is because the stage parallelism enables a larger word-length, even though the multiplexer delay limits the precision of each CSA stage in comparison to RCA. For OA, an even larger gap in word-length can be observed when compared to RCA. Additionally, unlike RCA and CSA of which the maximum word-length drops gradually with the increment of frequency, OA maintains full precision across a large range of frequencies. This is expected, because the critical path delay of OA is theoretically irrelevant to the operand word-length as discussed in Section~\ref{subsec:online_adder}. However, we also observe that given a large frequency requirement, the word-length of OA drops drastically. Besides, CSA cannot be implemented with larger precision than RCA either, because in this case the multiplexer delay becomes comparable to the delay of the carry chain for CSA, and it inhibits the benefits of parallelism. 
%
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.5\textwidth]{./figures/exp/area_adders.eps}
  \vspace{-4ex}
  \caption{The maximum word-length of different adder structures.}
  %\vspace{-2ex}
  \label{Fig:area_adder}
\end{figure}

@@

\section{Evaluation of Optimum Adder Structure}\label{sec:optimum_adder}
Generally, our evaluation could be of interest to a circuit designer in several ways. For example, suppose the algorithm designer wish the circuit to run at a given frequency within a specific area budget, while achieving the minimum output error. Besides, for many applications, the circuit designer would wish the circuit to operate as fast as possible with the minimum resource usage, whilst a certain error budget can be tolerated. In both cases, decisions must be made on which adder structure achieves this minimum, and which design scenario should be adopted.
%whether the circuit should be overclocked.

>>>>

\usepackage{algorithm}
\usepackage{array}
\usepackage{algpseudocode}
\usepackage{indentfirst}

\ifx\pdfoutput\undefined
\usepackage{graphicx}
\else
\usepackage[pdftex]{graphicx}
\fi
\usepackage{epstopdf}

%\usepackage[pdftex]{graphicx}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{multirow}

@@

In this paper, we provide detailed evaluations of design trade-offs for adders, which are key arithmetic primitives for a broad range of applications. In order to provide the optimum design choice for adders, we examine several design constrains and options, such as accuracy, operating frequency, silicon area, different types of adder structure, as well as different forms of computer arithmetic for adder implementation. In comparison to the conventional design scenario that timing closure is achieved by truncating the operand word-length, we take a radical step forward by adopting an alternative design scenario, which is named ``overclocking scenario''. In this case, circuits are operated far beyond the conservation region with full precision, while timing violation can be tolerated. Previous work has shown that for many applications, it is preferable to follow the overclocking scenario because the worst case is triggered only by certain input patterns and hence it happens rarely~\cite{SKfccm13}. In addition to the adder structures that are created based on standard binary arithmetic, such as ripple carry adder and carry select adder, we also evaluate adder with ``online arithmetic''~\cite{Ercegovac_OnlineOverview}, which employs different style of data representation to eliminate carry propagation. We demonstrate experimentally on FPGAs that for limited area budgets, applying the overclocking approach to ripple carry adders can be more beneficial than using fast adders to achieve similar latency. With more relaxed area constraints, the online adder outperforms in terms of accuracy or performance. Potentially our study will provide circuit designers with guidance and extra options when balancing between various design trade-offs.

@@

For an $n$-bit RCA, it is composed of $n$ serial-connected full adders (FA) as shown in Fig.~\ref{Fig:rca}. Typically the maximum frequency of RCA is determined by the longest carry propagation. Under the assumption that the carry propagation delay of each FA is a constant value $\mu_i=\mu$, the critical path of the RCA is: $\mu_{RCA}=n\mu$. For the sampling period $T_S$, it follows that if $T_S\geqslant\mu_{RCA}$, correct result will always be sampled. Otherwise, intermediate result will be sampled and potentially generating overclocking errors. The mean value of overclocking error is modeled~\cite{SKfccm13} as given by (\ref{Eq:MeanError_RCA}), where coefficient $b$ is given in (\ref{Eq:b}) and it determines the maximum length of error-free carry propagation.
%
\begin{eqnarray}\label{Eq:MeanError_RCA}
    E_{oc}=\left\{
        \begin{matrix}
            2^{-b}-2^{-n-1}, & \textrm{if $b\leq n$}\\
            0, & \textrm{otherwise}
        \end{matrix}
        \right.
\end{eqnarray}
\vspace{-1.5ex}
%
\begin{eqnarray}\label{Eq:b}
    b:=\left\lceil \frac{T_S}{\mu} \right\rceil=\left\lceil \frac{1}{\mu\cdot f_S}\right\rceil
\end{eqnarray}
\vspace{-1ex}
\normalsize

In the conventional scenario, a specific timing target can be met by truncating the word-length of RCA while timing violation is not permitted. This will result in truncation error for most data.  The mean value of truncation error is also modeled~\cite{SKfccm13} as given by~(\ref{Eq:MeanTruncError_RCA}), where parameters $k$ and $n$ denote the word-length of RCA before and after truncation, respectively.
%
\begin{eqnarray}\label{Eq:MeanTruncError_RCA}
    E_{trad}=\left\{
        \begin{matrix}
            2^{-n}-2^{-k}, & \textrm{if $n<k$}\\
            0, & \textrm{otherwise}
        \end{matrix}
        \right.
\end{eqnarray}
\vspace{-2ex}

@@

\subsection{Online Adder}\label{subsec:online_adder}
In addition to standard binary arithmetic, alternative forms of computer arithmetic were studied to boost performance. For example, adder with online arithmetic is also designed for low latency at the cost of extra silicon area. The structure of a digit-parallel online adder (OA) where all signals represented with SD numbers of digit set $\{-1,0,1\}$ is shown in Fig.~\ref{Fig:Radix2SD_adder}. The module ``3:2'' denotes a 3:2 compressor, which takes three inputs and generates two outputs, and is logically equivalent to a full adder (FA). A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As labeled in Fig.~\ref{Fig:Radix2SD_adder}, ideally the computation delay of this adder is only two FA delays for any operand word-length, at the expenses of one extra FA for each digit of operands.
%
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.45\textwidth]{./figures/SDAdder.eps}
  \vspace{-1ex}
  \caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
  \label{Fig:Radix2SD_adder}
  \vspace{-2ex}
\end{figure}

@@

As can be seen in Fig.~\ref{Fig:max_wl_adder}, for a relatively relaxed frequency requirement (e.g. 470 MHz), both CSA and OA can be implemented with greater word-lengths than RCA. For CSA, this is because the stage parallelism enables a larger word-length, even though the multiplexer delay limits the precision of each CSA stage in comparison to RCA. For OA, an even larger gap in word-length can be observed when compared to RCA. Additionally, unlike RCA and CSA of which the maximum word-length drops gradually with the increment of frequency, OA maintains full precision across a large range of frequencies. This is expected, because the critical path delay of OA is theoretically irrelevant to the operand word-length as discussed in Section~\ref{subsec:online_adder}. However, we also observe that given a large frequency requirement, the word-length of OA drops drastically. Besides, CSA cannot be implemented with larger precision than RCA either, because in this case the multiplexer delay becomes comparable to the delay of the carry chain for CSA, and it inhibits the benefits of parallelism.
%
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.5\textwidth]{./figures/exp/area_adders.eps}
  \vspace{-4ex}
  \caption{The maximum area of different adder structures.}
  %\vspace{-2ex}
  \label{Fig:area_adder}
\end{figure}

@@

\section{Evaluation of Optimum Adder Structure}\label{sec:optimum_adder}
Generally, our evaluation could be of interest to a circuit designer in several ways. For example, suppose the algorithm designer wish the circuit to run at a given frequency within a specific area budget, while achieving the minimum output error. Besides, for many applications, the circuit designer would wish the circuit to operate as fast as possible with the minimum resource usage, whilst a certain error budget can be tolerated. In both cases, decisions must be made on which adder structure achieves this minimum, and which design scenario should be adopted. In our experiments, the inputs are set to randomly generated data following uniform distribution.
%whether the circuit should be overclocked.
